Synthesizing Symmetric Lenses

### 2019-02-28 papers/SymmetricLenses/main.tex
minor changes

However, the set of real-world use cases for bijective lenses, where two
data formats contain different arrangements of precisely the same
information, is  
limited.  Oftentimes, two data
formats share just {\em some} of their information content.  For instance,
two related database tables might share four out of five columns, but each
might have a different fifth column.  Or one ad-hoc system-configuration
file might include some format-specific metadata, such as a date or a
reference number, while the same configuration file on another operating
system does not.  Indeed among the 48 benchmark problems described in
Section \ref{sec:evaluation} below, all of the benchmarks taken from Flash
Fill~\cite{flashfill} and many of the benchmarks that synchronize between
two ad hoc file formats have this characteristic (13 benchmarks in
total).  \bcp{The safer wording is better, but the reader is still left
  wondering ``If they are demonstrating a system for synthesizing
  non-bijective lenses, why did they choose a benchmark suite where 35/48
  examples are bijective??''}

Can Optician's basic synthesis procedure be extended to a richer class of
lenses? Could we even imagine synthesizing all \emph{symmetric
  lenses}~\cite{symmetric-lenses}---a much larger class that includes
bijective lenses, ``asymmetric'' lenses (where the transformation from a
source structure to a target can throw away information that must
then be restored when transferring from target to source), and even more flexibile
transformations that allow each side to retain information not present in
the other?

@@

This leaves us with the second difficulty in synthesizing symmetric
lenses: Whereas the number of bijective lenses between two given formats is
typically tiny, the number of simple symmetric lenses is typically enormous.
If a na\"ive search algorithm just selects the first simple symmetric lens
it finds, this will generally \emph{not} be the one the user wanted.  We
need a new principle for identifying ``more likely'' lenses and a more
sophisticated synthesis algorithm that uses this principle to search the
space more intelligently.

@@

With simple symmetric lenses and this SRE-based likelihood measure in hand, we
propose a new algorithm for synthesizing likely lenses. At its core, the
algorithm performs a type-directed search between the descriptions of the source
and target formats, measuring success using the likelihood measure.

Interesting complications arise from the need to deal with regular-expression
equivalences. There are infinitely many regular expressions equivalent to a
given one, and the lens returned by a type-directed search will in general
depend on which of the possible representations are chosen for its source and
target formats. \bcp{There's a disconnect here: The reader was probably not
  thinking about Boomerang's type-equivalence rule (indeed, very likely, did
  not even know there was such a thing).}Moreover, if Boomerang's type-equivalence rule is disallowed,
there may be {\em no} well-typed lens connecting a particular pair of source
and target formats 
selected by the user: in general, the synthesis algorithm has to search through
equivalent regular expression types to find the most likely lens. To tame
this complexity, we divide the synthesis algorithm into two communicating search
procedures (Figure~\ref{fig:high-level-algorithm}), following~\citet{optician}.
The first, \Expand, uses rewriting rules to propose new pairs of
stochastic regular expressions equivalent to the original pair. The second,
\GreedySynth, uses a greedy, type-directed
algorithm to find a simple symmetric lens between input SRE pairs, returning the lens and
its likelihood score to \Expand. The whole synthesis algorithm heuristically
terminates when a sufficiently likely lens is found.

@@

We added this synthesis procedure to the Boomerang system and explored its
effectiveness on a range of applications. Users set up a Boomerang synthesis
task by providing the source and target types and optionally supplying one or
more examples. In addition, users can override the default mechanism for
calculating the information content of a SRE by
asserting that certain strings are \emph{essential} or \emph{irrelevant},
thereby forcing certain data to either be retained or discarded during the
transformations.

We evaluated our system on a benchmark suite of 48 problems, including
data-cleaning tasks taken from Flash Fill~\cite{flashfill}, view-maintenance
tasks taken from Augeas~\cite{augeas2}, and ad-hoc synchronization tasks
from a variety of other sources (mostly from the bidirectional programming
literature).  \bcp{``... and we found...''?}  \bcp{Also, this paragraph is
  redundant with the corresponding contribution a bit below.  I think it
  would improve the flow if we made this one a bit longer (and stronger) and
  the contribution paragraph shorter, just summarizing what this one says.}

In summary, our contributions are as follows:
\begin{itemize}
%
\item We define {\em simple symmetric lenses}, a subclass of symmetric
lenses with no hidden state,
%
we propose a syntax for them---a collection
of combinators for building simple symmetric lenses---and we
describe how to calculate their likelihoods
(\S\ref{sec:overview}--\S\ref{sec:ssl}).  
\item We give an algorithm for synthesizing lenses built from these combinators,
  using a novel application of stochastic regular expressions to guide the
  search process (\S\ref{sec:synthesis}) and prove that our metrics have deep
  semantic meaning. \bcp{``Semantic meaning'' is redundant.  And ``deep
    semantic meaning'' just sounds silly.  Can we say it more precisely?}
%
\item We extend the Boomerang implementation with simple symmetric lenses
and lens synthesis specifications, 
  including types, examples, and relevance annotations. We evaluate our
  implementation, the effects of
  optimizations, and our inclusion of relevance annotations on a set of
  48 lens synthesis benchmarks drawn from data cleaning, view maintenance, and
  file synchronization tasks. The system can synthesize simple symmetric lenses
  for all of the benchmarks in under 30 seconds (\S\ref{sec:evaluation}).
%
\item We prove that simple symmetric lenses fall between asymmetric
lenses~\cite{Focal2005-long} and full symmetric
lenses~\cite{symmetric-lenses} in expressiveness.  Indeed, the
class of simple lenses can be characterized semantically as the subset of
full symmetric lenses for which a simple ``forgetfulness'' property holds
(\S\ref{sec:related}).
\end{itemize}
Along the
way, we show that {\em star-semiring equivalences}\bcp{citation} on regular
expressions 
remain valid when extended to stochastic regular expressions, preserving
probability distributions as well as languages (\S\ref{sec:sre}).
We close with related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).
Elided proofs can be found in the appendices
\ifappendices\else included with the supplemental material\fi.

@@

\subsection{Simple Symmetric Lenses}
Semantically, a simple symmetric lens (we will just say ``lens'' when clear
from context) between 
sets $X$ and $Y$ comprises four functions subject to four round-tripping
laws.
\vspace*{.7ex}
\begin{trivlist}
\item 
%\begin{enumerate}
%\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
%\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
%\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
%\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
%\end{enumerate}
\begin{center}
\begin{minipage}[c]{.3\textwidth}
\begin{gather}
 \mbox{$\CreateR \OfType X \rightarrow Y$} \nonumber \\
 \mbox{$\CreateL \OfType Y \rightarrow X$} \nonumber \\
 \mbox{$\PutR \OfType X \rightarrow Y \rightarrow Y$} \nonumber \\
 \mbox{$\PutL \OfType Y \rightarrow X \rightarrow X$} \nonumber
\end{gather} 
\end{minipage} 
\hspace{.4in}
\begin{minipage}[c]{.5\textwidth}
\begin{align}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{align}
\end{minipage} 
\end{center}
\end{trivlist}
The two \KW{create} functions are used to \proposedelete{propagate shared
  information and} fill 
in default values when introducing new data (\EG, on \KW{create} the ``unk''
salary entry is added alongside the name to the management file when HR inserts
a new employee). The two \KW{put} functions propagate edits from one format to
the other by combining a new value from one with an old value from the other.
The record projection notation $\Lens.\PutR$ extracts the \PutR function
from the 
lens $\Lens$. These four functions can be used to keep the
common information between two file types in sync. For example, if a new file
of the left-hand format is created, the \CreateR function will build a new
file in the 
right-hand format. If the right-hand file is then edited, the \PutL function can
update the left-hand file with the changed information.

The round-tripping laws guarantee that pushing an ``un-edited'' value from one
side (the result of a put or create) back through a lens in the other direction
will get back to exactly where we began. For example, consider a lens
synchronizing the employee data formats in
Figure~\ref{fig:minimized-representations}. Applying the \CreateR function to a
salary file creates a insurance file with the same employee names in the file,
with
\lstinline{"UNK"} for each insurance company. The round-tripping
laws guarantees that applying the \PutL function to the generated insurance file
and the original salary file will return the original salary file, unmodified. If \KW{put}ting a string $s$ into a string $t$ results in the original string $s$, then $s$ and $t$ are considered \emph{synchronized}.

Simple symmetric lenses differ from ''classical'' symmetric lenses in that
they do not involve a \emph{complement}. We give a detailed comparison in
\S\ref{sec:relationship}.

@@

In principle, programmers can define lenses just by writing
four arbitrary functions by hand and showing they satisfy the round-tripping
laws, but this can be tedious and error prone.  A better idea is to provide
a set of combinators---a domain-specific language---for building complex,
law-abiding 
lenses from simpler ones. Such bidirectional languages have been given for
many different kinds of data, including database
relations~\cite{BohannonPierceVaughan} algebraic data
types~\cite{symmetric-lenses}, graphs~\cite{tgg}, and more.
We focus here on combinators for defining lenses between string data formats
described by regular expressions.

If $\BRegex$ and $\BRegexAlt$ are regular expressions, then
\lstinline{$\Lens$ : $\BRegex \Leftrightarrow \BRegexAlt$} indicates that
$\Lens$ is a simple symmetric lens between $\LanguageOf{\BRegex}$ and
$\LanguageOf{\BRegexAlt}$. (We will use undecorated variables later for
stochastic regular expressions, so we mark plain REs with overbars.)
We illustrate some of these
combinators by defining lenses on subcomponents of the employee data formats.

The simplest combinator is the identity lens \IdentityLens, which takes as an
argument a regular expression $\BRegex$ and propagates data unchanged
in both directions.
%
\begin{lstlisting}
id(name) : name $\Leftrightarrow$ name
\end{lstlisting}
%
The identity lens moves data back and forth from source to target without
changing it. Both the \CreateR and \CreateL functions are the identity function
($\CreateR \App s = s$), and the put functions merely return the first argument
($\PutROf{\String}{\StringAlt} = \String$).

In contrast, $\DisconnectOf{\Regex}{\RegexAlt}{\String}{\StringAlt}$ does not propagate any data at all
from one format to the other.  The \Disconnect lens takes four arguments: two
regular expressions ($\Regex, \RegexAlt$) and two strings ($\String, \StringAlt$). The regular expressions specify the formats
on the two sides, while the strings provide default values. 
%
\begin{lstlisting}
disconnect(salary, "", "unk", "") : salary $\Leftrightarrow$ ""
\end{lstlisting}
%
On creates, the input values are thrown away, and default values are
returned ($\CreateLOf{\StringAlt} =
\lstinline{"unk"}$), and on puts, the second argument is used and the first is
thrown away ($\PutROf{\String}{\StringAlt} = \StringAlt$). For 
example, the \lstinline{salary} field is only present in management files,
so the disconnect lens can ensure salary edits do not cause updates to the
HR file.
With the above lens, $\PutL \App \lstinline{""} \App 60000$ will return 60000,
and \PutR will always return \lstinline{""}.

@@

Finally, there are a number of combinators that construct
more complex lenses from simpler ones. These include concatenation
($\ConcatLensOf{\Lens_1}{\Lens_2}$, often written $\Lens_1 . \Lens_2$), variation
($\OrLensOf{\Lens_1}{\Lens_2}$), and iteration ($\IterateLensOf{\Lens}$).
For example, we could use the lens
%
\begin{lstlisting}
id(name) . id(" ") . id(name) . del(": ") . ins(",")
\end{lstlisting}
%
to transform ``\lstinline{Jane Doe: }'' to ``\lstinline{Jane Doe,}'' in
the left-to-right direction.

% concat(ins(header),employees_lens) : emp_salaries $\Leftrightarrow$ emp_insurance
%
%\ksf{I don't think this example is necessary.}
%Data can also appear in multiple forms. For example,
%a list of employees with insurance information can either be empty or
%be a sequence of employee/insurance records: \ksf{This example is
%vacuous as you could eliminate the or in favor of the *.  Do somethign
%else to introduce or?}
%
%\begin{lstlisting}
%or(id(""),employees_lens) : emp_salaries $\Leftrightarrow$ "" | ("\n" emp_ins) . ("\n" emp_ins)*
%\end{lstlisting}
The \IterateLens lens is useful for synchronizing a series of items or rows in a
table. For example given a
lens \lstinline{employee_lens} that synchronizes data for a single employee, the
lens\bcp{In all these examples, I'd like to have one more space on each side
of the ``:''}
%
\begin{lstlisting}
iterate(id("\n") . employee_lens) : ("\n" . emp_salary)$^*$ $\Leftrightarrow$ ("\n" . emp_ins)$^*$
\end{lstlisting}
%
transforms a list of employees in employees in the Management format to a list
of employees in the HR format and vice versa. These combinators
are combined in figure~\ref{fig:example_lens} to construct a complete lens
between the employee formats.

@@

\bcp{I wonder whether the rest of this subsection can be cut... It is both
  hard to understand and too brief to really say much, and I wonder whether
  we need it for anything between here and where we give it a real
  explanation.}

Most of the typing rules for lenses are syntax directed, where only a handful of
rules result in a given type\bcp{syntax directed usually means that the
  conclusion of at most
  {\em one} rule will match a given goal; if that's not the case here, then
  we need to speak more carefully}.  For example, the type of a
\ConcatLens lens is the concatenation of the regular expressions of the sub-lenses.
\[
  \inferrule*
  {
    \Lens_1 \OfType \BRegex_1 \Leftrightarrow \BRegexAlt_1\\
    \Lens_2 \OfType \BRegex_2 \Leftrightarrow \BRegexAlt_2
  }
  {
    \ConcatLensOf{\Lens_1}{\Lens_2} \OfType \BRegex_1 \Concat \BRegex_2
    \Leftrightarrow
    \BRegexAlt_1 \Concat \BRegexAlt_2
  }
\]
The typing  rules for maintain structural similarity
between regular expression types and lens terms, with one exception: The
type equivalence rule permits a well-typed lens to be well-typed among any
equivalent regular expression pair.
%
\[
  \centering
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \BRegex \SSREquiv \BRegex'\\
    \BRegexAlt \SSREquiv \BRegexAlt'
  }
  {
    \Lens \OfType \BRegex' \Leftrightarrow \BRegexAlt'
  }
\]
%
This rule is needed for lenses like the final one in
Figure~\ref{fig:example_lens} to be well-typed.

@@

For larger and more complex formats, it can become quite difficult to
write lenses by hand, as we did above.  Instead, we would like to synthesize
them from types and examples.  
That is, given a pair of regular expressions and a set of input-output
examples, we want 
to find a simple symmetric lens that is typed by the regular expression pair and
satisfies all the input/output examples. We this a \emph{satisfying
  lens}. For our running example, suppose we wish to synthesize a lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance}, using as sole
input-output example the data in Figure~\ref{fig:minimized-representations}.

One challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries, emp_insurance, ..., ...)
\end{lstlisting}
%
is another well-typed lens that satisfies the example in
Figure~\ref{fig:minimized-representations} (where \lstinline{...} represents
components of the example itself\bcp{huh?}). In general, {\em many} examples
may be 
required to rule out all possible occurrences of nested\bcp{??} \Disconnect lenses,
particularly in complex formats.
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
\proposedelete{While we could ask programmers to provide detailed and precise specifications
for their synthesis tasks, we would prefer an algorithm that can intelligently
choose a lens despite the underspecification.} Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

How can we identify such a ``likely'' lens? We propose the following heuristic:
A satisfying lens is ``more likely'' if it uses more data from one format to
construct the other.  For example, the
identity lens (which uses all the data) is more likely than the disconnect
lens (which uses none). Formally, we define the {\em likelihood} of a
satisfying lens as the 
expected number of bits required to recover data in one format from 
data in the other; higher likelihoods correspond to fewer bits. Two
strings $s$ and $t$ are \emph{synchronized according to lens
  $l$}\bcp{Earlier we used $\Lens$, now we are switching to $l$...} if $l.\PutR\
s\ t = t$ and $l.\PutL\ t\ s = s$. We can \emph{recover} $s$ from 
$t$ using bits $b$ and lens $l$ if we can reconstruct $s$ from $t$, $b$, and
$l$. For example, given the \IdentityLens{} lens, we can recover $s$ from $t$
using no bits because, in this case, $s$ is just $t$. In contrast, given the
\Disconnect{} lens, we need enough bits to fully encode $s$ in order to recover
it from $t$ because all the information in $t$ gets thrown away. Thus, if both
\IdentityLens{} and \Disconnect{} are satisfying lenses for a particular pair of
formats, \IdentityLens{} will be more likely.

@@

\begin{theorem}
  If $\Regex \SSREquiv \RegexAlt$ then $\ProbabilityOf{\Regex}{\String} =
  \ProbabilityOf{\RegexAlt}{\String}$, for all strings $\String \in \LanguageOf{\Regex}$.
\end{theorem}

(For the proof, see the appendix \ifappendices
Theorem~\ref{thm:semantics-correctness} \else{} included with the
supplemental material\fi.)  This theorem will come in handy as we traverse
(with \Expand) or normalize across (as part of \GreedySynth) star-semiring
equivalences.

@@

Intuitively, these equations state that complements are uniquely determined by
the most recently input $x$ and $y$. Such lenses correspond exactly with simple
symmetric lenses, where all state is maintained by the $x$ and $y$ data.

>>>>

However, the set of real-world use cases for bijective lenses, where two
data formats contain different arrangements of precisely the same
information, is  
limited.  Oftentimes, two data
formats share just {\em some} of their information content.  For instance,
one ad-hoc system-configuration
file might include some format-specific metadata, such as a date or a
reference number, while the same configuration file on another operating
system does not.  Indeed among the 48 benchmark problems described in
Section \ref{sec:evaluation} below, all of the benchmarks taken from Flash
Fill~\cite{flashfill} and many of the benchmarks that synchronize between
two ad hoc file formats have this characteristic.

Can Optician's basic synthesis procedure be extended to a richer class of
lenses? Could we even imagine synthesizing all \emph{symmetric
  lenses}~\cite{symmetric-lenses}---a much larger class that includes
bijective lenses, ``asymmetric'' lenses (where the transformation from a
source structure to a target can throw away information that must
then be restored when transferring from target to source), and even more flexibile
transformations that allow each side to throw away information?

@@

This leaves us with the second difficulty in synthesizing symmetric
lenses: Whereas the number of bijective lenses between two given formats is
typically tiny, the number of simple symmetric lenses is typically enormous.
If a na\"ive search algorithm just selects the first simple symmetric lens
it finds, the returned lens will generally \emph{not} be the one the user wanted.  We
need a new principle for identifying ``more likely'' lenses and a more
sophisticated synthesis algorithm that uses this principle to search the
space more intelligently.

@@

With simple symmetric lenses and this SRE-based likelihood measure in hand, we
propose a new algorithm for synthesizing likely lenses. At its core, the
algorithm performs a type-directed search between descriptions of the data
formats, measuring success using the likelihood measure.

Interesting complications arise from the need to deal with regular-expression
equivalences. There are infinitely many regular expressions equivalent to a
given one, and the lens returned by a type-directed search will in general
depend on which of the possible representations are chosen for its source and
target formats. Moreover, certain lenses may not be well-typed unless the
format representations are replaced by equivalent ones:
in general, the synthesis algorithm has to search through
equivalent regular expression types to find the most likely lens. To tame
this complexity, we divide the synthesis algorithm into two communicating search
procedures (Figure~\ref{fig:high-level-algorithm}), following~\citet{optician}.
The first, \Expand, uses rewriting rules to propose new pairs of
stochastic regular expressions equivalent to the original pair. The second,
\GreedySynth, uses a greedy, type-directed
algorithm to find a simple symmetric lens between input SRE pairs, returning the lens and
its likelihood score to \Expand. The whole synthesis algorithm heuristically
terminates when a sufficiently likely lens is found.

@@

We added this synthesis procedure to the Boomerang system and explored its
effectiveness on a range of applications. Users set up a Boomerang synthesis
task by providing two regular expressions and optionally supplying input-output
examples. Users can also override the default mechanism for
calculating the information content of a SRE by
asserting that certain strings are \emph{essential} or \emph{irrelevant},
forcing certain data to either be retained or discarded during the
transformations.

We evaluate our implementation, the effects of optimizations, and our inclusion
of relevance annotations on a set of 48 lens synthesis benchmarks drawn from
data cleaning, view maintenance, and file synchronization tasks. We find the
system can synthesize simple symmetric lenses for all of the benchmarks in under
30 seconds (\S\ref{sec:evaluation}).

In summary, our contributions are:
\begin{itemize}
%
\item We define {\em simple symmetric lenses}, a subclass of symmetric
lenses that do not maintain any hidden state (\S\ref{sec:overview}).  
%
We present a collection of combinators for building simple symmetric lenses and
describe how to calculate their likelihoods (\S\ref{sec:ssl}).
\item We give an algorithm for synthesizing lenses built from these combinators,
  using a novel application of stochastic regular expressions to guide the
  search process (\S\ref{sec:synthesis}) and prove that our metrics have deep
  semantic meaning.
%
\item We extend the Boomerang implementation with simple symmetric lenses,
  demonstrating that they integrate smoothly with other lens extensions such as
  quotient lenses~\cite{quotientlenses} and matching
  lenses~\cite{matchinglenses}.

\item We also extend Boomerang with symmetric lens synthesis specifications:
  including types, examples, and relevance annotations. We evaluate our system
  on a set of 48 benchmarks, and find it synthesizes each of them in under
  30 seconds.
%
\item We show that simple symmetric lenses fall between asymmetric
lenses~\cite{Focal2005-long} and full symmetric
lenses~\cite{symmetric-lenses} in expressiveness.  Indeed, the
class of simple lenses can be characterized semantically as the subset of
full symmetric lenses for which a simple ``forgetfulness'' property holds
(\S\ref{sec:related}).
\end{itemize}
Along the
way, we show the {\em star-semiring equivalences} on regular expressions
remain valid when extended to stochastic regular expressions, preserving
probability distributions as well as languages (\S\ref{sec:sre}).
We close with related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).
Elided proofs can be found in the appendices
\ifappendices\else included with the supplemental material\fi.

@@

\subsection{Simple Symmetric Lenses}
Semantically, a simple symmetric lens (or lens when clear from context) between
sets $X$ and $Y$ comprises four functions subject to four ``round-tripping''
laws.

%\begin{enumerate}
%\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
%\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
%\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
%\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
%\end{enumerate}
\begin{center}
\begin{minipage}[c]{.3\textwidth}
\begin{gather}
 \mbox{$\CreateR \OfType X \rightarrow Y$} \nonumber \\
 \mbox{$\CreateL \OfType Y \rightarrow X$} \nonumber \\
 \mbox{$\PutR \OfType X \rightarrow Y \rightarrow Y$} \nonumber \\
 \mbox{$\PutL \OfType Y \rightarrow X \rightarrow X$} \nonumber
\end{gather} 
\end{minipage} 
\hspace{.4in}
\begin{minipage}[c]{.5\textwidth}
\begin{align}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{align}
\end{minipage} 
\end{center}
\noindent
\vskip -.5em
The two \KW{create} functions are used to propagate shared information and fill
in default values when introducing new data (\EG, on \KW{create} the ``unk''
salary entry is added alongside the name to the management file when HR inserts
a new employee). The two \KW{put} functions propagate edits from one format to
the other by combining a new value from one with an old value from the other.
The notation $\Lens.\PutR$ extracts the \PutR function from the
lens $\Lens$; similarly for \PutL. These four functions can be used to keep the
common information between two file types in synch. For example, if a new file
of the left-hand format is created, the \CreateR function will a new file in the
right-hand format. If the right-hand format is edited, the \PutL function can
update the left-hand format with the changed information.

The round-tripping laws guarantee that pushing an ``un-edited'' value from one
side (the result of a put or create) back through a lens in the other direction
will get back to exactly where we began. For example, consider a simple
symmetric lens
synchronizing the employee data formats in
Figure~\ref{fig:minimized-representations}. Applying the \CreateR function to a
salary file creates a insurance file with the same employee names in the file,
with
\lstinline{"UNK"} for each insurance company. The round-tripping
laws guarantees that applying the \PutL function to the generated insurance file
and the original salary file will return the original salary file, unmodified. If \KW{put}ting a string $s$ into a string $t$ results in the original string $s$, then $s$ and $t$ are considered \emph{synchronized}.

This simple symmetric lens definition differs from classical symmetric lenses in
that it does not involve a \emph{complement}. We give a detailed comparison in
\S\ref{sec:relationship}.

@@

In principle, programmers can define simple symmetric lenses just by writing
four arbitrary functions by hand and showing they satisfy the round-tripping
laws. But a better idea is to streamline the process by providing a set of
combinators---a domain-specific language---for building complex, law-abiding
lenses from simpler ones. Such bidirectional languages have been given for
building lenses over many different kinds of data, including database
relations~\cite{BohannonPierceVaughan} algebraic data
types~\cite{symmetric-lenses}, graphs~\cite{tgg}, and more.

We focus here on combinators for defining lenses between string data formats
described by regular expressions.
If $\BRegex$ and $\BRegexAlt$ are regular expressions, then
\lstinline{$\Lens$ : $\BRegex \Leftrightarrow \BRegexAlt$} indicates that
$\Lens$ is a simple symmetric lens between $\LanguageOf{\BRegex}$ and
$\LanguageOf{\BRegexAlt}$. (We will use undecorated variables later for
stochastic regular expressions; we mark plain REs with overbars.)
We illustrate some of these
combinators by defining lenses on subcomponents of the employee data formats.

The simplest combinator is the identity lens \IdentityLens, which takes as an
argument a regular expression $\BRegex$ and performs an identity mapping on data
matching $\BRegex$. 
%
\begin{lstlisting}
id(name) : name $\Leftrightarrow$ name
\end{lstlisting}
%
The identity lens moves data back and forth from source to target without
changing it. Both the \CreateR and \CreateL functions are the identity function
($\CreateR \App s = s$), and the put functions merely return the first argument
($\PutROf{\String}{\StringAlt} = \String$).

In contrast, $\DisconnectOf{\Regex}{\RegexAlt}{\String}{\StringAlt}$ does not propagate any data at all
from one format to the other.  The \Disconnect lens takes four arguments: two
regular expressions ($\Regex, \RegexAlt$) and two strings ($\String, \StringAlt$). The regular expressions specify the formats
on the two sides, while the strings provide default values. 
%
\begin{lstlisting}
disconnect(salary, "", "unk", "") : salary $\Leftrightarrow$ ""
\end{lstlisting}
%
On creates, the input values are thrown away, and default values are
returned ($\CreateLOf{\StringAlt} =
\lstinline{"unk"}$), and on puts, the second argument is used and the first is
thrown away ($\PutROf{\String}{\StringAlt} = \StringAlt$). For 
example, the \lstinline{salary} field is only present in management files,
so the disconnect lens can ensure salary edits do not cause updates to the
HR file.
For example, with the above lens, $\PutL \App \lstinline{""} \App 60000$ will return 60000,
and \PutR will always return \lstinline{""}.

@@

Finally, there are a number of combinators that construct
more complex lenses from simpler ones. These include concatenation
($\ConcatLensOf{\Lens_1}{\Lens_2}$, often written $\Lens_1 . \Lens_2$), variation
($\OrLensOf{\Lens_1}{\Lens_2}$), and iteration ($\IterateLensOf{\Lens}$).
For example, we could use the following lens
%
\begin{lstlisting}
id(name) . id(" ") . id(name) . del(": ") . ins(",")
\end{lstlisting}
%
to transform ``\lstinline{Jane Doe: }'' to ``\lstinline{Jane Doe,}'' in
the left-to-right direction.

% concat(ins(header),employees_lens) : emp_salaries $\Leftrightarrow$ emp_insurance
%
%\ksf{I don't think this example is necessary.}
%Data can also appear in multiple forms. For example,
%a list of employees with insurance information can either be empty or
%be a sequence of employee/insurance records: \ksf{This example is
%vacuous as you could eliminate the or in favor of the *.  Do somethign
%else to introduce or?}
%
%\begin{lstlisting}
%or(id(""),employees_lens) : emp_salaries $\Leftrightarrow$ "" | ("\n" emp_ins) . ("\n" emp_ins)*
%\end{lstlisting}
The \IterateLens lens is useful for synchronizing a series of items or rows in a
table. For example given a
lens \lstinline{employee_lens} that synchronizes data for a single employee the
following lens
%
\begin{lstlisting}
iterate(id("\n") . employee_lens) : ("\n" . emp_salary)$^*$ $\Leftrightarrow$ ("\n" . emp_ins)$^*$
\end{lstlisting}
%
transforms a list of employees in employees in the Management format to a list
of employees in the HR format in the left-to-right direction. These combinators
are combined in figure~\ref{fig:example_lens} to construct a complete lens
between the employee formats.

@@

Most of the typing rules for lenses are syntax driven, where only a handful of
rules result in a given type.  For example, the type of a
\ConcatLens lens is the concatenation of the regular expressions of the sub-lenses.
\[
  \inferrule*
  {
    \Lens_1 \OfType \BRegex_1 \Leftrightarrow \BRegexAlt_1\\
    \Lens_2 \OfType \BRegex_2 \Leftrightarrow \BRegexAlt_2
  }
  {
    \ConcatLensOf{\Lens_1}{\Lens_2} \OfType \BRegex_1 \Concat \BRegex_2
    \Leftrightarrow
    \BRegexAlt_1 \Concat \BRegexAlt_2
  }
\]
With one notable exception, lens typing rules maintain structural similarity
between regular expression types and lens terms; type equivalence does not. Type
equivalence permits a well-typed lens to be well-typed among any equivalent
regular expression pair.
%
\[
  \centering
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \BRegex \SSREquiv \BRegex'\\
    \BRegexAlt \SSREquiv \BRegexAlt'
  }
  {
    \Lens \OfType \BRegex' \Leftrightarrow \BRegexAlt'
  }
\]
%
This rule is needed for lenses like the final one in
Figure~\ref{fig:example_lens} to be well-typed.

@@

Even for small examples, lenses like the one above take a little getting used
to. For larger and more complex formats, it can become quite difficult to
write lenses by hand.  Instead, we wish to synthesize them from examples. 
That is, given a pair of regular expressions and a set of input-output examples, we want
to find a simple symmetric lens that is typed by the regular expression pair and
satisfies all the input/output examples. We call such a lens a \emph{satisfying
  lens}. In our running example, we wish to synthesize a lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance}, using as an input-output
example the data in Figure~\ref{fig:minimized-representations}.

One challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries, emp_insurance, ..., ...)
\end{lstlisting}
%
is another well-typed lens that satisfies the example in
Figure~\ref{fig:minimized-representations} (where \lstinline{...} represents
components of the example itself). In general, {\em many} examples may be
required to rule out all possible occurrences of nested \Disconnect lenses,
particularly in complex formats.
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
While we could ask programmers to provide detailed and precise specifications
for their synthesis tasks, we would prefer an algorithm that can intelligently
choose a lens despite the underspecification. Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

How can we identify such a ``likely'' lens? We propose the following heuristic:
A satisfying lens is ``more likely'' if it uses more data from one format to
construct the other. One manifestation of this principle involves choosing the
identity lens (which uses all the data) rather than the disconnect lens (which
uses none of the data). We define the likelihood of a satisfying lens as the
expected number of bits required to recover data in one format from synchronized
data in the other. Higher likelihoods correspond to fewer bits. Two
strings $s$ and $t$ are \emph{synchronized according to lens $l$} if $l.\PutR\
s\ t = t$ and $l.\PutL\ t\ s = s$. We can \emph{recover} $s$ from
$t$ using bits $b$ and lens $l$ if we can reconstruct $s$ from $t$, $b$, and
$l$. For example, given the \IdentityLens{} lens, we can recover $s$ from $t$
using no bits because, in this case, $s$ is just $t$. In contrast, given the
\Disconnect{} lens, we need enough bits to fully encode $s$ in order to recover
it from $t$ because $t$ contains no useful information. Thus, if both
\IdentityLens{} and \Disconnect{} are satisfying lenses for a particular pair of
formats, then \IdentityLens{} will be more likely.

@@

\begin{theorem}
  If $\Regex \SSREquiv \RegexAlt$ then $\ProbabilityOf{\Regex}{\String} =
  \ProbabilityOf{\RegexAlt}{\String}$, for all strings $\String \in \LanguageOf{\Regex}$.
\end{theorem}
\begin{proof}
  See the appendix %
\ifappendices  Theorem~\ref{thm:semantics-correctness} \else{} included with the supplemental
material\fi.
\end{proof}

This theorem will come in handy as we traverse (with \Expand) or normalize
across (as part of \GreedySynth) star-semiring equivalences.

@@

Intuitively, these equations state that complements are uniquely determined by
the most recent input $x$ and $y$. Such lenses correspond exactly with simple
symmetric lenses, where all state is maintained by the $x$ and $y$ data.
